Staged Abstract Interpreters

### 2018-11-17 paper/22_bg_abs.tex
revise

Based on the generic interface we presented in Section~\ref{bg_lang}, 
now we describe the instantiation of an abstract interpreter.
To keep the simplicity, we intend to use simply abstract domains, and just establish a 
context-/path-/flow-insensitive, store-widened analysis in this section --- indeed, it is coarse
but simple enough to setup a foundation for the staged abstract interpreter we will present later.
However, in Section~\ref{cfa}, we will see how to regain context-/path-/flow-sensitivity 
for pushdown control-flow analysis; in Section~\ref{cases_imp}, we will see how to instantiate 
an interval abstract domain for numerical analysis.

\todo{maybe put to related work} other big-step abstract interpreter, 
big-step abstract interpreter using delimited control operators \cite{Wei:2018:RAA:3243631.3236800},
monadic big-step abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17},
arrow-based abstract interpreter \cite{Keidel:2018:CSP:3243631.3236767}.

\paragraph{Abstract Instantiation.}
Our instantiation roughly follows the Abstracting Abstract Machines methodology 
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the concrete semantics 
to abstract semantics. The store now maps addresses to sets of abstract values, meaning that an address points
to all possible values may occur at runtime, where the abstract value is either a closure or 
a single abstract number \texttt{NumV}, which stands for the top. The address space is also
constrained to be finite, in the monovariant setting, we simply use variable names for it.

@@

\subsection{Stage Polymorphic Lattices} \label{stagedpoly_lat}
To effectively reuse the code between unstaged and staged part, we also make the lattice structure to
be stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

For example, here we use powerset as the abstract domain, and the unstaged lattice 
of powerset can be easily implemented as follows:

@@

The ordering relation is to ask whether one set is a subset of the other, 
join is to union the two set, and meet is to intersect the two sets.
Accordingly, other structures used in the rest of the paper such as tuples 
(products) and maps can be lifted to lattices element-wise or point-wise.
The code for them are elided.

\subsection{Abstract Semantics}
The operations @get@ and @put@ on abstract stores are slightly changed according to
the abstract semantics: @get@ uses the bottom element in the @Value@ lattice if the 
queried address not exists; @put@ performs the update by joining with the existing values.
@alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to our abstract domain, i.e., 
a singleton set that contains only one \texttt{CloV} or \texttt{NumV} object.

\todo{conditional, use different store?}
\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (fs: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- fs) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$ $\sqcup$ $\sigma$_*; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = 
    thn $\sqcup$ els
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    Set(NumV())
} // to be continued
\end{lstlisting}

For branching, @branch0@ simply joins the answer from two branches. For arithmetics, @prim_eval@ simply 
returns the top abstract number element.
It is worth to note that @apply_closure@ is where we handle the non-determinism of applications.
The non-determinism happens because of the abstract store maps addresses to sets of abstract values;
when dereferencing an address from the store, we need to explore all possible values to achieve a sound 
analysis. In the case of @apply_closure@, the first argument @fs@ may contains multiple target closures,
so we have a @for@ comprehension over all closures, apply them respectively and form a set result values @vs@;
meanwhile, every time when we evaluate into the body expression @e@ we use the latest store $\sigma$@_*@
which is updated iteratively. Thus it is a store-widened analysis.

\subsection{Fixpoint Iteration}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also widely used in the name of co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL}.

The idea is to set up two caches called @in@ and @out@, which are both mapping from the arguments of interpreter 
(@Config@) to the returned value of interpreter (@Ans@). The @in@ contains what we already computed from the last
iteration, the @out@ cache is what we computed after this iteration joined with the previous one.
When starting a new iteration, the @in@ is set to be the result of last iteration, i.e., @out@; the @out@ is set
to be empty. In the iteration, we first ask whether @out@ contains what we want, if yes then it is returned;
otherwise, we retrieve what we have from @in@, compute the result for this time, and put the joined result back to @out@.
Note that we also instrument the recursive call by putting @cached_ev@ to @evev@.
After one iteration, if @in == out@, then there is no more information can be gained, thus the iteration should 
end and we have reached the fixed point.

>>>>

Based on the generic interface we presented in Section~\ref{bg_lang}, 
we now describe the instantiation of an abstract interpreter.
To keep the presentation simple, we use simple abstract domains, and just establish a 
context-/path-/flow-insensitive, store-widened analysis in this section --- indeed, it is coarse
but simple enough to setup a foundation for the staged abstract interpreter we will present later.
In Section~\ref{cfa}, we will see how to regain context-/path-/flow-sensitivity 
for pushdown control-flow analysis; in Section~\ref{cases_imp}, we will see how to instantiate 
an interval abstract domain for numerical analysis.

\paragraph{Abstract Instantiation.}
Our instantiation roughly follows the Abstracting Abstract Machines methodology 
\cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12} that transforms the concrete semantics 
to abstract semantics. The store now maps addresses to sets of abstract values, meaning that an address points
to all possible values that may occur at runtime, where the abstract value is either a closure or 
a single abstract number \texttt{NumV}, which stands for the top element of the number lattice. 
The address space is also constrained to be finite, in the monovariant setting, we simply use variable names 
for address.

@@

\subsection{Stage Polymorphic Lattices} \label{stagedpoly_lat}
To effectively reuse the code between unstaged and staged variants, we also make the lattice structure
stage polymorphic. A lattice type class parameterizes over an element type \texttt{E} and defines
five operations for \texttt{R[E]}, such as meet, join, and ordering relation; again, \texttt{R} 
is the higher-kinded type indicating the binding-time.
\begin{lstlisting}
trait Lattice[E, R[+_]] {
  val bot: R[E];  val top: R[E]
  def $\sqsubseteq$(l1: R[E], l2: R[E]): R[Boolean]
  def $\sqcup$(l1: R[E], l2: R[E]): R[E]
  def $\sqcap$(l1: R[E], l2: R[E]): R[E]
}
\end{lstlisting}

For example, here we use powersets as the abstract domain, and the unstaged lattice 
of powersets can be easily implemented as follows:

@@

The ordering relation is to ask whether one set is a subset of the other, 
join is to union the two sets, and meet is to intersect the two sets.
Accordingly, other structures used in the rest of the paper such as tuples 
(products) and maps can be lifted to lattices element-wise or point-wise.
The code for them are elided.

\subsection{Abstract Semantics}
The operations @get@ and @put@ on abstract stores are slightly changed according to
the abstract semantics: @get@ uses the bottom element in the @Value@ lattice if the 
queried address does not exist; @put@ performs the update by joining with the existing values.
function @alloc@ simply invokes the @Addr@ constructor.
\texttt{close} and \texttt{num} lift syntactic literals to our abstract domain, i.e., 
a singleton set that contains only one \texttt{CloV} or \texttt{NumV} object.

\begin{lstlisting}
object AbsInterp extends Abstract {
  type R[+T] = T
  val $\rho$0 = Map[Ident, Addr]();  val $\sigma$0 = Map[Addr, Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = 
    $\sigma$.getOrElse(a, Lattice[Value].bot)
  def put($\sigma$: Store, a: Addr, v: Value): Store =
    $\sigma$ + (a -> (v $\sqcup$ get($\sigma$, a)))
  def alloc($\sigma$: Store, x: Ident): Addr = Addr(x)
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = Set(CloV($\lambda$, $\rho$))
  def num(i: Lit): Value = Set(NumV())
  def apply_closure(ev: EvalFun)
    (fs: Value, arg: Value, $\sigma$: Store): Ans = { var $\sigma$_* = $\sigma$
      val vs = for (CloV(Lam(x, e), $\rho$) <- fs) yield {
        val $\alpha$ = alloc($\sigma$_*, x)
        val (v, v$\sigma$) = ev(e, put($\rho$, x, $\alpha$), put($\sigma$_*, $\alpha$, arg))
        $\sigma$_* = v$\sigma$ $\sqcup$ $\sigma$_*; v
      }
    (vs.reduce(Lattice[Value].$\sqcup$), $\sigma$_*)
  }
  def branch0(cnd: Value, thn: =>Ans, els: =>Ans): Ans = 
    thn $\sqcup$ els
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    Set(NumV())
} // to be continued
\end{lstlisting}

For branching, @branch0@ joins the answers from two branches since we are doing a path-insensitive analysis. 
For arithmetics, @prim_eval@ returns the top abstract number element.
It is worth noting that @apply_closure@ is where we handle the non-determinism of applications.
The non-determinism happens because the abstract store maps addresses to sets of abstract values;
when dereferencing an address from the store, we need to explore all possible values to achieve a sound 
analysis. In the case of @apply_closure@, the first argument @fs@ may contains multiple target closures,
so we use a @for@ comprehension over all closures, apply them respectively and form a set result values @vs@;
meanwhile, every time when evaluating the body expression @e@ we use the latest store $\sigma$@_*@
which is updated iteratively during the loop. Thus this is a store-widened analysis.

\subsection{Fixpoint Iteration}
We described the abstract semantics modularly in the last section, however, the abstract 
interpreter may not terminate for some input program. In this section, we use a memoization
technique to ensure its termination. This technique is also widely used under the name of co-inductive caching algorithm
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800} or truncated depth-first evaluation 
\cite{Rosendahl:AbsIntPL}.

The idea is to set up two caches called @in@ and @out@, which are both mapping from the arguments of the interpreter 
(@Config@) to the result values of the interpreter (@Ans@). The @in@ cache contains what we already have 
computed from the last
iteration, the @out@ cache is what we will have computed after this iteration joined with the previous one.
When starting a new iteration, @in@ is set to be the result of the last iteration, i.e., @out@; @out@ is set
to be empty. In the iteration, we first check whether @out@ contains what we want, if yes then it is returned;
otherwise, we retrieve what we have from @in@, compute the result for this iteration, 
and put the joined result back into @out@.
Note that we also instrument the recursive call by putting @cached_ev@ into @evev@.
After one iteration, if @in == out@, then there is no more information to be gained, thus the iteration should 
end and we have reached the fixed point.

### 2018-11-17 paper/23_bg_stage.tex
revise

\subsection{Multi-Stage Programming in LMS}
Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a multi-stage programming framework implemented 
as a Scala library that enables dynamic code generation in a type-safe manner.
Different from the syntactic approach such as MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} that uses quotations, 
LMS distinguishes binding-time solely based on types.
LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary type, indicating an expression will be 
known at next stage. All operations acting on a @Rep@ expression will be resulted in the generated code. However,
we need to provide intermediate representation and code generation support for type @T@ to LMS. 
Fortunately, the LMS framework already provides
such support for primitive types and commonly used data structure such as arrays and maps; implementing such support
for custom class is also straightforward. Let's go back and see how LMS can be used to specialize the power function
we mentioned in Section 1.

@@

In the code shown above, \texttt{power} takes 2 arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as well, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and the body of this function multiples @b@ five times.

\subsection{Staged Concrete Semantics} 
In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations.
Meanwhile, @RepConcInterpOps@ also extends from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures. \todo{what are static, what are dyanmic?}

@@

The first notable change is the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. For @num@, we use @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage.
The interesting thing is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, which
is a pair of the syntactic lambda term and the enclosing environment.
However, what we desire is a \textit{compiled} closure, instead of a lifted @CloV@ value like 
what we did for @NumV@ --- in other words, the syntactic term of the lambda expression should be 
eliminated and specialized away.
The specialization of the interpreter with respect to the body of the lambda term proceed under the
assumptions that the argument and latest store will be provided later. \note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an abstract interpreter modularly.
In the end of @close@, we create a @CompiledClo@ object that contains a staged function, which all will be
exhibited in the generated code.

@@

Then, @apply_closure@ takes three arguments the function value, argument value and the latest store, 
which are both of @Rep@ type. It means basically there is nothing we can do at current stage --- 
so we create a new node @ApplyClosure@ that contains these arguments in the intermediate representation graph,
and in the code generation phase we can emit the code that does the application. \todo{reflectEffect}.

@@

\paragraph{Code Generation.}
When generating code for the next stage, values like @CompiledClo@ and @NumV@ are kept as what they are because
they are intended to be existed in the next stage; however, we need to treat @ApplyClosure@ a little bit 
differently.

@@

Concretely, as shown in the code above, when an @ApplyClosure@ is matched when emitting code, we just 
emit a function application where the function value is extracted from the @CompiledClo@
(we assume the object program is well-typed so that we can safely cast it to @CompiledClo@), 
and argument and store are the rest fields in @ApplyClosure@.

>>>>

\subsection{Multi-Stage Programming in LMS}
Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a multi-stage programming framework implemented 
as a Scala library that enables dynamic code generation in a type-safe manner.
Different from the syntactic approach in MetaOCaml \cite{DBLP:conf/flops/Kiselyov14} that uses quotations, 
LMS distinguishes binding-time solely based on types.
LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary type, indicating an expression will be 
known at next stage. All operations acting on a @Rep[T]@ expression will be residualized in the generated code. However,
we need to provide intermediate representation and code generation support for type @T@ to LMS. 
Fortunately, the LMS framework already provides
such support for primitive types and commonly used data structure such as arrays and maps; implementing such support
for custom class is also straightforward. Let's go back and see how LMS can be used to specialize the power function
we mentioned in Section 1.

@@

In the code shown above, \texttt{power} takes two arguments where \texttt{b} is declared as \texttt{Rep[Int]} type 
meaning that \texttt{b} is a representation of \texttt{Int} but whose value will be available at the next stage. 
Meanwhile, the result of the @power@ function will be available at the next stage as well, thus the return value 
of \texttt{power} is also of type \texttt{Rep[Int]}. 
Then we specialize @power@ in function @snippet@ by providing @x@ to @5@.
The generated code for \texttt{power5} is a function that takes only one argument which corresponds to @b@,
and the body of this function multiples @b@ five times.

\subsection{Staged Concrete Semantics} 
In the staging part, we share the same concrete type instantiation, 
i.e. trait \texttt{Concrete}, but we reimplement the staged version for the concrete operations
in trait @RepConcInterpOps@.
Meanwhile, it is also extended from @LMSOps@ which provides necessary supports for the 
staged version of primitive types and data structures.
To stage the interpreter, we first identifies the syntactic input is known statically, 
and the interpreter returns a pair of values and store, which both becomes dynamic values.
Thus the staged stores propogate the binding-time to environments, because they share the same
address component.

@@

The first notable change is that the abstract type member @R@ is assigned to be @Rep@, and accordingly the affected 
types such as @Env@ and @Store@ become @Rep[Env]@ and @Rep[Store]@. For @num@, we use the @lift@ function which is a
built-in in LMS that lifts a current-stage value to next stage.
The interesting point is how we handle closures and function applications when staging is involved.
In the unstaged version, function @close@ is used to lift a literal lambda term to a closure, which
is a pair of the syntactic lambda term and the enclosing environment.
However, what we desire is a \textit{compiled} closure, instead of a lifted @CloV@ value like 
what we did for @NumV@ --- in other words, the syntactic term of the lambda expression should be 
eliminated and specialized away.
The specialization of the interpreter with respect to the body of the lambda term proceeds under the
assumptions that the argument and latest store will be provided later. 
%\note{to denotational style?, back to HO?}
We will see this is an important observation that enables us to specialize an abstract interpreter in a modular way.
At the end of function @close@, we create a @CompiledClo@ object that contains a staged function, which all will be
emitted in the generated code.

@@

Next, @apply_closure@ takes three arguments: the function value, argument value and the latest store, 
which are both of @Rep@ type. This means basically there is nothing we can do at the current stage --- 
so we create a new node @ApplyClosure@ that contains these arguments in the intermediate representation graph,
and in the code generation phase we can emit the code that performs the application. The additional @reflectEffect@
is a function in LMS that handles side effects.

@@

\paragraph{Code Generation.}
When generating code for the next stage, values like @CompiledClo@ and @NumV@ are kept as they are because
they are intended exist in the next stage; however, we need to treat @ApplyClosure@ a little differently.

@@

Concretely, as shown in the code above, when emitting code for @ApplyClosure@, we just 
emit a function application where the function value is extracted from the @CompiledClo@ object
(we assume the input program is well-typed so that we can safely cast it to @CompiledClo@).
Argument and store are the remaining rest fields in @ApplyClosure@.
